\chapter{\abstractname}

Analyzing a function's running time is crucial in the analysis of algorithms.
Timing functions are an essential step to measure the exact running time of a function.
With their help, we can proof a certain running time class.
In the proof assistant Isabelle, the conversion step of a function to their running time equivalent is done manually.
The manual step introduces errors and makes following proofs worthless.
This work aims to implement an automatic converter for functions to their timing functions in Isabelle.

Here, we consider the number of function calls as a measurement of the running time of a function.
We analyze existing translation schemas for non-curried first-order functions.
One provides an extension to support higher-order and curried functions.
Implementing the schema for curried functions is not possible due to strict typing in Isabelle.
Therefore, we restrict the converter's schema to non-curried higher-order functions.
We formalize a correctness proof for the translation schema for first-order functions.

The implemented converter simplifies the conversion of functions.
It offers automatic translations through a single command and proves termination automatically.
We use a tactic based on the termination proof of the original function.
This tactic covers most functions.
Manual termination is also possible in case of failure.
We discuss the current restrictions of the converter, including functions within datatypes and dealing with functions without cost.
